Data Specialization for a Staged Lambda Calculus

### 2014-11-11 paper/figures/splitting-overview.tex
more minor fixes
a bit of work on terminology section

\begin{figure}
\textbf{Languages:}
\begin{itemize}
\item \lang: two-staged lambda calculus with product, sum, and
  recursive types
\item \langmono: an unstaged lambda calculus with products.
\end{itemize}

\textbf{\lang\ Evaluation Relations:}
\begin{itemize}
\item 
\bbone-Evaluation: $e\mathbin{\redonesym}[\xi;v]$, where $\xi$ is a \emph{residual table} and $v$ is a \emph{partial value}.

@@

\item 
Table reification: \ur{Fill in...}
\end{itemize}

% $red_2$: 2-eval,  which we think of as being identical to specialization.

\vspace{.75em}
\textbf{\lang\ Splitting:}

@@

\hspace{4em}$c$ is the \emph{combined term} (representing all stage~\bbone\ subcomputations in $e$)

\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputations in $e$)

@@

\hspace{4em}$\maskt{\xi;v}$ and \texttt{(let l=b in r)} reduce (via $\redsym$) to identical values.

@@

\hspace{4em}$p$ is the \emph{precomputation} (representing all stage~\bbone\ subcomputations in $e$)

\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputations in $e$)

\hspace{2em}\bbtwo-Splitting Correctness: If $e~\redtwosym~q$, and $q~\redsym~v$, then \texttt{(let l=b in r)}~$\redsym~v$

\caption{Summary of \lang\ evaluation and splitting.}
\label{fig:terminology}
\label{fig:termSplitSummary}
\end{figure}

>>>>

\begin{figure*}
\textbf{Languages:}
\begin{itemize}
\item \lang: two-staged typed lambda calculus with product, sum, and
  recursive types
\item \langTwo: monostaged typed lambda calculus
\item \langmono: an unstaged lambda calculus
\end{itemize}

\textbf{\lang\ Dynamic Semantics:}
\begin{itemize}
\item 
\bbone-Evaluation: $e\mathbin{\redonesym}[\xi;v]$, where $\xi$ is a \emph{residual table} and $v$ is a \emph{partial value}.

@@

\item 
Table reification: $[\xi;q]\mathbin{\reifysym} q'$, where $q,q'$ are residuals,
$\xi$ is a residual table.

\item
Residual evaluation: $q\mathbin{\redsym} v$, where $q$ is a residual, $v$ is a
value.
\end{itemize}

\textbf{\lang\ Splitting:}

@@

\hspace{4em}$c$ is the \emph{combined term}
(representing all stage~\bbone\ subcomputations in $e$)
in \langmono

\hspace{4em}$l.r$ is the \emph{resumer}
(representing all stage~\bbtwo\ subcomputations in $e$)
in \langmono

@@

\hspace{4em}$\maskt{\xi;v}$ and \texttt{(let l=b in r)} 
reduce (via $\redsym$) to equivalent values.

@@

\hspace{4em}$p$ is the \emph{precomputation}
(representing all stage~\bbone\ subcomputations in $e$)
in \langmono

\hspace{4em}$l.r$ is the \emph{resumer}
(representing all stage~\bbtwo\ subcomputations in $e$)
in \langmono

\hspace{2em}\bbtwo-Splitting Correctness:
If $e~\redtwosym~q$, and $q~\redsym~v$, then
\texttt{(let l=p in r)}~$\redsym~v'$ and $v,v'$ are equivalent.
% TODO: FIX THIS
% if e reduces to q, then p reduces to u
% and q == [u/l]r
% that is, stage 1 reduces, even if stage 2 doesn't

\caption{Summary of \lang\ evaluation and splitting.}
\label{fig:termSplitSummary}
\end{figure*}

### 2014-11-11 paper/obsolete/correctness.tex
a bit of work on terminology section

>>>>

\subsection{Metatheory}

Recall that residuals live in \langTwo; we will indicate typing judgments in
\langTwo\ with $\vdash_\bbtwo$.

%\begin{definition}
%Context $\Gamma$ is well-formed ($\Gamma\wf$) if it
%contains only stage-2 variables.
%\end{definition}

\begin{definition}
An environment $\xi$ is well-formed ($\Gamma\vdash\xi\wf$) if either:
\begin{enumerate}
\item $\xi = \cdot$; or
\item $\xi = \xi',x:B\mapsto e$ where
$\Gamma\vdash\xi'\wf$ and
$\typeslangTwo[\Gamma,\dom{\xi'}] e B$
%$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
%$\Gamma,\dom{\xi'}\vdash e \res$.
\end{enumerate}
\end{definition}

\begin{theorem}
If $\typeswor e A$ then $\Gamma\wf$ and $A\istypewor$.
\end{theorem}

\begin{theorem}
If $\diaonesub$ and $\typesone e A$ then
\begin{enumerate}
\item $\Gamma\vdash\xi\wf$;
\item $\Gamma,\dom\xi\vdash \colone{v}{A}$; and
\item $\Gamma,\dom\xi\vdash v\pval$.
\end{enumerate}
\end{theorem}

\begin{theorem}
If $\diatwosub$ and $\typestwo e A$ then
\begin{enumerate}
\item $\typeslangTwo q A$; and
\item $\Gamma\vdash_\bbtwo q\val$.
\end{enumerate}
\end{theorem}

\begin{theorem}\label{thm:reify-type}
If $\Gamma\vdash\xi\wf$ and
$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
then 
$\reify{\xi}{\hat y}{q}$ and
$\typeslangTwo q A$.
\end{theorem}

\TODO
Note somewhere how to run stage-one non-$\fut A$ terms. For example, a stage-one
integer term is guaranteed not to depend on the table, although one might be
produced. One may either discard the table, or evaluate everything in the table
(and terminating with the partial value iff everything in the table terminates).

### 2014-11-11 paper/semantics.tex
improve figures, yet again
a bit of work on terminology section

We present an overview of the formal development along with our
terminology, as summarized in \ref{fig:termSplitSummary}.

\ref{fig:termSplitSummary}

Our starting point is a two-staged lambda calculus, whose syntax and
types are very similar to the multi-stage languages considered in the
literature on partial evaluation and meta programming.  Specifically,
we consider a two-stage, which we call \lang\, which is variant of
Davies`s multi-stage language.  Apart from the restriction to two
stages instead of multiple stages, there are no significant
differences between our syntax and the static semantics and those of
Davies'. We restrict our attention to two stages, because it
simplifies the (many) details of the splitting problem, without
altering the essence of the problem.  We expect that the techniques
presented here can be extended to multi-stage languages.

The dynamic semantics of \lang\ differs from Davies's approach and
similar approaches in meta programming literature, because it avoids
duplication of staged work. Duplication takes place in meta
programming when a second-stage computation is silently duplicated by
the semantics.  In addition to leading to increases in run time, a
duplicating semantics makes it harder to reason about the splitting
algorithm, whose development and exposition is our main goal.  Apart
from the non-duplication, the dynamic semantics of \lang\ is
essentially identical to  Davies'.

@@

\input {figures/grammars}
\input {figures/statics}

@@

@@

\crem{Cut references to type validity.}
The type validity judgment $\Delta \vdash A \istypewor$, defined in
\cref{fig:validTypes}, ensures that a type $A$ exists at stage $w$; for
example, any type $\fut A$ only exists at stage \bbone.  The typing judgment
$\typeswor e A$ says that $e$ has type $A$ at stage $w$, in the context
$\Gamma$, and is defined in \cref{fig:statics}.  As one would expect, the
typing judgment only produces valid types.
%(by induction on the derivation of the typing judgment).
% under the assumptions in $\Delta$. This context is only augmented by recursive types.

@@

As is the case in essentially all staged language, the dynamic
semantics evaluates a term to a residual (a term in the target
language \langmono) in two stages. In the first stage, all stage-1
terms are evaluated by a relation, called {\em 1-evaluation}; in the
second stage all stage-2 terms are evaluated by a relation
$\redtwosym$, called {\em 2-evaluation}.  To eliminate duplication,
our semantics lifts out each stage-2 term nested inside of a stage-1
expressions, into a table which makes it possible for the stage-2
term---more precisely the corresponding residual obtained by reducing
all stage-1 subterms---to be referred to by use a dynamically
generated variable (a label).

@@

Since stage-1 evaluation does not evaluate stage-2 terms and since
stage-2 terms may be nested inside stage-1 terms, 1-evaluation cannot
reduce \lang\ terms to values.  Instead 1-evaluation yields a {\em
  partial value}, which which is a stage~\bbone\ term that have been
evaluated except for its stage-2 subterms, which can only
be~\bbtwo\ variables wrapped in $\next$ blocks.  \ref{fig:validPvalues}
defines partial values.

@@

The relation 1-evaluation, defined by $\redonesym$ judgment as shown
in \ref{fig:diaSemantics}, takes an stage \bbone\ term%
\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
the type system ensures that those must occur under a $\next$. Consequently
$\redonesym$ will never directly encounter a variable.}
%
to a residual table $\xi$ and a partial value $v$.  For non-staging
features of \lang, $\redonesym$ is essentially standard call-by-value
evaluation but in addition gathers the subterms' residual tables into
a single table. \ur{The syntax of residual tables should be defined in
  the terminology figure} Residual table entries are created by the
evaluation of \next: 1-evaluation reduces a \next\ expression by
reducing the body of the \next\ with 2-evaluation to obtain a
residual $q$ and returns a residual table that maps a freshly
generated variable to the residual.

@@

%% Umut: superseded by what is above.
%% Version Nov 7, 2014
%% \section{\texorpdfstring{\lang}{λ12} Statics and Dynamics}
%% \label{sec:semantics}

%% \input {figures/grammars}
%% \input {figures/statics}

%% \lang\ is a typed two-stage lambda calculus with products, sums, and
%% isorecursive types.  Its statics are adapted from the linear temporal
%% logic given by \cite{davies96}, restricted to two stages and extended
%% with general sums and recursion.%
%% \footnote{This restriction to two stages is made primarily for simplicity of
%% presentation. All of these techniques could feasibly be extended to more
%% complicated stage systems.} 
%% Although \cite{davies96} was mostly interested in describing partial evaluation,
%% the system is equally apt as an input to splitting.

%% \subsection{Statics}

%% In \lang, the $\next$ constructor includes a stage \bbtwo\ term in a
%% stage~\bbone\ term. To ensure that the staging features are only applied to
%% valid terms, we index our judgments not only by a type, but also a stage $w$.

%% The type validity judgment $\Delta \vdash A \istypewor$, defined in
%% \cref{fig:validTypes}, ensures that a type $A$ exists at stage $w$; for
%% example, any type $\fut A$ only exists at stage \bbone.  The typing judgment
%% $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in the context
%% $\Gamma$, and is defined in \cref{fig:statics}.  As one would expect, the
%% typing judgment only produces valid types.
%% %(by induction on the derivation of the typing judgment).
%% % under the assumptions in $\Delta$. This context is only augmented by recursive types.

%% Except for the staging features $\next$, $\prev$, and $\pause$, the ordinary
%% language features (i.e., the introductory and eliminatory forms for product,
%% sum, and function types) preserve the stage of their subterms, and work at both
%% stages \bbone\ and \bbtwo. Thus, if we were to remove the staging features,
%% \lang\ would simply consist of two non-interacting copies of a standard lambda
%% calculus. Variables in the context are annotated with the stage at which they
%% were introduced and can only be used at the same stage.

%% Note that there is no way to eliminate a term of type $\fut A$ into a stage
%% \bbone\ term---the only elimination form is $\prev$, which results in a stage
%% \bbtwo\ term. This is why well-typed terms cannot have any information flow
%% from stage \bbtwo\ to stage \bbone, and what makes splitting possible.

@@

%% \crem{Can we cut \ref{fig:diaSemanticsSpec} from the paper? Perhaps just put a
%% few representative rules into \ref{fig:diaSemantics}?}

@@

%\begin{remark}
%For any $\colone{e}{A}$ containing no $\next$ subexpressions, $\redonesym$ will
%always compute an empty environment, and a partial value identical to the result
%of call-by-value evaluation of $e$.
%%derivationally equivalent to standard call-by-value evaluation.
%\end{remark}

%\subsection{Metatheory}
%
%Recall that residuals live in \langTwo; we will indicate typing judgments in
%\langTwo\ with $\vdash_\bbtwo$.
%
%%\begin{definition}
%%Context $\Gamma$ is well-formed ($\Gamma\wf$) if it
%%contains only stage-2 variables.
%%\end{definition}
%
%\begin{definition}
%An environment $\xi$ is well-formed ($\Gamma\vdash\xi\wf$) if either:
%\begin{enumerate}
%\item $\xi = \cdot$; or
%\item $\xi = \xi',x:B\mapsto e$ where
%$\Gamma\vdash\xi'\wf$ and
%$\typeslangTwo[\Gamma,\dom{\xi'}] e B$
%%$\Gamma,\dom{\xi'}\vdash \coltwo{e}{B}$ and
%%$\Gamma,\dom{\xi'}\vdash e \res$.
%\end{enumerate}
%\end{definition}
%
%\begin{theorem}
%If $\typeswor e A$ then $\Gamma\wf$ and $A\istypewor$.
%\end{theorem}
%
%\begin{theorem}
%If $\diaonesub$ and $\typesone e A$ then
%\begin{enumerate}
%\item $\Gamma\vdash\xi\wf$;
%\item $\Gamma,\dom\xi\vdash \colone{v}{A}$; and
%\item $\Gamma,\dom\xi\vdash v\pval$.
%\end{enumerate}
%\end{theorem}
%
%\begin{theorem}
%If $\diatwosub$ and $\typestwo e A$ then
%\begin{enumerate}
%\item $\typeslangTwo q A$; and
%\item $\Gamma\vdash_\bbtwo q\val$.
%\end{enumerate}
%\end{theorem}
%
%\begin{theorem}\label{thm:reify-type}
%If $\Gamma\vdash\xi\wf$ and
%$\Gamma,\dom\xi\vdash \colone{\next\ \hat y}{\fut A}$
%then 
%$\reify{\xi}{\hat y}{q}$ and
%$\typeslangTwo q A$.
%\end{theorem}

%\TODO
%Note somewhere how to run stage-one non-$\fut A$ terms. For example, a stage-one
%integer term is guaranteed not to depend on the table, although one might be
%produced. One may either discard the table, or evaluate everything in the table
%(and terminating with the partial value iff everything in the table terminates).

>>>>

%We present an overview of the formal development along with our
%terminology, as summarized in \ref{fig:termSplitSummary}.

Our starting point is \lang, a typed, two-stage lambda calculus with products,
sums, and isorecursive types. This language is adapted from \cite{davies96}, and
is very similar to the staged languages considered in the partial evaluation and
metaprogramming literature. In this paper, we restrict our attention to a
two-stage language to simplify our presentation of the splitting algorithm, but
we expect that the techniques presented here extend to more stages.

The dynamic semantics of \lang\ differs from Davies's approach and
similar approaches in metaprogramming literature, because it avoids
duplication of staged work. Duplication takes place in meta
programming when a second-stage computation is silently duplicated by
the semantics.  In addition to leading to increases in run time, a
duplicating semantics makes it harder to reason about the splitting
algorithm, whose development and exposition is our main goal.  Apart
from the non-duplication, the dynamic semantics of \lang\ is
essentially identical to  Davies'.

@@

\crem{Fold previous section in here.}

\input {figures/grammars}

@@

\input {figures/statics}

@@

The typing judgment $\typeswor e A$ says that $e$ has type $A$ at stage $w$, in
the context $\Gamma$, and is defined in \cref{fig:statics}.

\crem{Notice that certain types are impossible.}

@@

As is the case in essentially all staged languages, the dynamic
semantics evaluate a term to a residual (a term in the target
language \langTwo) in two stages. In the first stage, all stage-1
terms are evaluated by a relation, called {\em 1-evaluation}; in the
second stage all stage-2 terms are evaluated by a relation
$\redtwosym$, called {\em 2-evaluation}.  To eliminate duplication,
our semantics lifts out each stage-2 term nested inside of a stage-1
expressions, into a table which makes it possible for the stage-2
term---more precisely the corresponding residual obtained by reducing
all stage-1 subterms---to be referred to by use a dynamically
generated variable (a label).

@@

Since stage-1 evaluation does not evaluate stage-2 terms and since
stage-2 terms may be nested inside stage-1 terms, 1-evaluation cannot
reduce \lang\ terms to values.  Instead 1-evaluation yields a {\em
  partial value}, which which is a stage~\bbone\ term that have been
evaluated except for its stage-2 subterms, which can only
be~\bbtwo\ variables wrapped in $\next$ blocks.  \ref{fig:grammar}
defines partial values.

@@

The relation 1-evaluation, defined by $\redonesym$ judgment as shown
in \ref{fig:diaSemantics}, takes an stage \bbone\ term%
\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
the type system ensures that those must occur under a $\next$. Consequently
$\redonesym$ will never directly encounter a variable.}
%
to a residual table $\xi$ and a partial value $v$.  For non-staging
features of \lang, $\redonesym$ is essentially standard call-by-value
evaluation but in addition gathers the subterms' residual tables into
a single table. Residual table entries are created by the
evaluation of \next: 1-evaluation reduces a \next\ expression by
reducing the body of the \next\ with 2-evaluation to obtain a
residual $q$ and returns a residual table that maps a freshly
generated variable to the residual.

@@

%% Umut: superseded by what is above.
%% Version Nov 7, 2014

@@

@@
